---
title: "Lab 2"
format: html
author: "Lucy Gagnon"
editor: visual
---

## Task 3

Write a function that takes as its inputs that data you simulated (or any data of the same type) and a sequence of $\theta$ values of length 1000 and produces Likelihood values based on the Binomial Likelihood. Plot your sequence and its corresponding Likelihood function.

The likelihood function is given below. Since this is a probability and is only valid over the interval from $[0, 1]$ we generate a sequence over that interval of length 1000.

You have a rough sketch of what you should do for this part of the assignment. Try this out in lab on your own.

```{r}

### Bernoulli LH Function ###
# Input: obs_data, theta
# Output: bernoulli likelihood
bernoulli_lh <- function(obs_data, theta) {
  n<-length(obs_data). # define num samples
  x<-sum(obs_data). # define sample data
  return(theta^x*(1-theta)^(n-x)) 
}


### Plot LH for a grid of theta values ###
# Create the grid #
thetas = seq(0, 1, length.out = 1000)
# Store the LH values
lh_values<- bernoulli_lh(obs_data, thetas)
# Create the Plot
plot(thetas, lh_values, type = "l", main= "Likelihood Profile", 
     xlab="Simulated support",ylab = "Likelihood")
```

## Task 4

Write a function that takes as its inputs prior parameters \textsf{a} and \textsf{b} for the Beta-Bernoulli model and the observed data, and produces the posterior parameters you need for the model. \textbf{Generate and print} the posterior parameters for a non-informative prior such as \textsf{(a,b) = (1,1)} and for an informative case \textsf{(a,b) = (3,1)}}.

```{r}

```
