---
title: "HW2"
format: pdf
author: "Lucy Gagnon"
editor: visual
---

```{r}
set.seed(123)
```

# Question 1

## Task 3

*Write a function that takes as its inputs that data you simulated (or any data of the same type) and a sequence of* $\theta$ *values of length 1000 and produces likelihood values based on the Binomial Likelihood. Plot your sequence and its corresponding Likelihood function.*

*The likelihood function is given below. Since this is a probability and is only valid over the interval from* $[0, 1]$ *we generate a sequence over that interval of length 1000.*

*You have a rough sketch of what you should do for this part of the assignment. Try this out in lab on your own.*

```{r, echo=TRUE}
# define observed data
obs_data <- rbinom(n=100, size=1, prob=0.01)
### Bernoulli LH Function ###
# Input: obs_data, theta
# Output: bernoulli likelihood
bernoulli_lh <- function(obs_data, theta) {
  n<-length(obs_data) # define num samples
  x<-sum(obs_data) # define sample data sum
  return(theta^x*(1-theta)^(n-x)) 
}


### Plot LH for a grid of theta values ###
# Create the grid #
thetas = seq(0, 1, length.out = 1000)
# Store the LH values
lh_values<- bernoulli_lh(obs_data, thetas)
# Create the Plot
plot(thetas, lh_values, type = "l", main= "Likelihood Profile", 
     xlab="Simulated support",ylab = "Likelihood")
```

## Task 4

*Write a function that takes as its inputs prior parameters* $\text{a}$ *and* $\text{b}$ *for the Beta-Bernoulli model and the observed data, and produces the posterior parameters you need for the model.* $\textbf{Generate and print}$ *the posterior parameters for a non-informative prior such as* $\text{(a,b)} = (1,1)$ *and for an informative case* $\text{(a,b)} = (3,1)$*.*

```{r}
# function to update prior params for Beta-Bernoulli
## inputs: observed data, a, b
update_params <- function(obs_data, a, b) {
  n <- length(obs_data) # define num samples
  x <- sum(obs_data) # define sample data sum

  results <- list(a_param = x + a, b_param = n - x + b) # append updated params
  return(results)
}
# define non-informative a,b
a1<-1
b1<-1
# define informative a,b
a2<-3
b2<-1

# run function for non-info case
output1 <- update_params(obs_data, a1, b1)
# extract a,b
a1_new<- output1$a_param
b1_new<- output1$b_param
# print
print(a1_new)
print(b1_new)

# run function for info case
output2 <- update_params(obs_data, a2, b2)
# extract a,b
a2_new<- output2$a_param
b2_new<- output2$b_param
# print
print(a2_new)
print(b2_new)
```

Here we wrote a function *update_params* that takes as its inputs prior parameters $\text{a}$ and $\text{b}$ for the Beta-Bernoulli model and the observed data, *obs_data*, and produces the posterior parameters we need for the model. The posterior parameters for a non-informative prior $(a,b) = (1,1)$ were $(a_n,b_n) = (2,100)$, and the posterior parameters for an informative case $(a,b) = (3,1)$ were $(a_n,b_n) = (4,100)$.

## Task 5

*Create two plots, one for the informative and one for the non-informative case to show the posterior distribution and superimpose the prior distributions on each along with the likelihood. What do you see? Remember to turn the y-axis off using the command* $yaxt=\text{"none"}$ *in the plot() command. Why? Superimposing the distributions may make the scale non-sense.*

```{r}
# Non-informative case
## define prior
prior1 <- dbeta(thetas, shape1 = a1, shape2 = b1)
## define posterior using updated params
posterior1 <- dbeta(thetas, shape1 = a1_new, shape2 = b1_new)

## plot the likelihood
plot(thetas, lh_values,type = "l", col = "blue", pch = 16,
     xlab = "Simulated support",ylab = "Likelihood",
     main = "Non-informative prior", yaxt = "n",
     xlim = range(thetas), yaxt="none")
# overlay prior
par(new = TRUE)
# plot prior
plot(thetas, prior1,type = "l", col = "red", pch = 17, lty = 2,
     axes = FALSE, xlab = "", ylab = "",xlim = range(thetas),
    yaxt="none")
# overlay posterior
par(new = TRUE)
# plot posterior
plot(thetas, posterior1,type = "l", col = "green", pch = 17, lty = 3,
     axes = FALSE, xlab = "", ylab = "",xlim = range(thetas),yaxt="none")
# add legend
legend("topright",legend = c("Likelihood", "Prior", "Posterior"),
       col = c("blue", "red", "green"), lty = c(1, 2, 3), pch = c(16, 17, 17))

# Informative Case
# define prior
prior2 <- dbeta(thetas, shape1 = a2, shape2 = b2)
# define posterior with updated parameters
posterior2 <- dbeta(thetas, shape1 = a2_new, shape2 = b2_new)

# plot the likelihood
plot(thetas, lh_values,type = "l", col = "blue", pch = 16,
     xlab = "Simulated support",ylab = "Likelihood",main = "Informative prior",
     yaxt = "n",xlim = range(thetas),yaxt="none")
# overlay prior
par(new = TRUE)
# plot prior
plot(thetas, prior2, type = "l", col = "red", pch = 17, lty = 2,
     axes = FALSE, xlab = "", ylab = "", xlim = range(thetas), yaxt="none")
# overlay posterior
par(new = TRUE)
# plot posterior
plot(thetas, posterior2, type = "l", col = "green", pch = 17, lty = 3,
     axes = FALSE, xlab = "", ylab = "", xlim = range(thetas),yaxt="none")
# add legend
legend("topright",legend = c("Likelihood", "Prior", "Posterior"),
       col = c("blue", "red", "green"),lty = c(1, 2, 3),pch = c(16, 17, 17))
### Here, I used ChatGPT to help with plot formatting/styles
### comments are my own
```

Here, we see the posterior overlap the likelihood in the case of the non-informative prior. When we implement an informative prior, the posterior moves away from the likelihood and approaches the prior.

## Question 2

## Part A

*Derive (step by step) the posterior density,* $p(\theta|x_{1:n})$*. Give the form of the posterior in terms of one of the most common distributions (Bernoulli, Beta, Exponential, or Gamma).*

*Proof.* Suppose we have data $x_1, \dots, x_n$ which wee are modeling as i.i.d. observations from an Exponential distribution with unknown parameter $\theta$. We write the likelihood as

$$
p(x_i \mid \theta)
=
\theta \exp(-\theta x_i),
\qquad x_i>0.
$$

Alsp, suppose the prior distribution is

$$
\theta \sim \operatorname{Gamma}(a,b),
$$

where $a$ is the shape parameter and $a$ is the rate parameter. The prior density is

$$
p(\theta)
=
\frac{b^a}{\Gamma(a)}
\theta^{a-1}
\exp(-b\theta),
\qquad \theta>0.
$$

We will find the posterior density $p(\theta \mid x_{1:n})$. By Bayes' theorem,

$$
p(\theta \mid x_{1:n})
=
\frac{p(\theta, x_{1:n})}{p(x_{1:n})}
=
\frac{
p(x_{1:n} \mid \theta)p(\theta)
}{
p(x_{1:n})
}.
$$

Because $p(x_{1:n})$ does not depend on $\theta$, it is absrobed into the constant of proportionality:

$$
p(\theta \mid x_{1:n})
\propto
p(x_{1:n} \mid \theta)p(\theta).
$$

The likelihood is given by

$$
\begin{aligned}
p(x_{1:n} \mid \theta)
&=
\prod_{i=1}^n p(x_i \mid \theta) \\
&=
\prod_{i=1}^n
\left[
\theta \exp(-\theta x_i)
\right] \\
&=
\theta^n
\exp\left(
-\theta\sum_{i=1}^n x_i
\right).
\end{aligned}
$$

Combining the likelihood and prior gives

$$
\begin{aligned}
p(\theta \mid x_{1:n})
&\propto
\left[
\theta^n
\exp\left(
-\theta\sum_{i=1}^n x_i
\right)
\right]
\left[
\frac{b^a}{\Gamma(a)}
\theta^{a-1}
\exp(-b\theta)
\right] \\
&\propto
\theta^{a+n-1}
\exp\left[
-\theta
\left(
b+\sum_{i=1}^n x_i
\right)
\right].
\end{aligned}
$$

This is the kernel of a Gamma distribution, with shape parameter $a+n$ and rate parameter $b+\sum_{i=1}^n x_i$. Therefore,

$$
\boxed{
\theta \mid x_{1:n}
\sim
\operatorname{Gamma}
\left(
a+n,\;
b+\sum_{i=1}^n x_i
\right).
}
$$

Thus, the posterior density is

$$
p(\theta \mid x_{1:n})
=
\frac{
\left(b+\sum_{i=1}^n x_i\right)^{a+n}
}{
\Gamma(a+n)
}
\theta^{a+n-1}
\exp\left[
-\theta
\left(
b+\sum_{i=1}^n x_i
\right)
\right],
\qquad \theta>0.
$$

QED

## Part B

*Why is the posterior distribution a proper density or probability distribution function?*

The posterior distribution is a proper density/probability distribution function because it takes the form of an updated Gamma distribution, where $a' = a+n >0$ and $b'=b+\sum_{i=1}^n x_i > 0$. The Gamma distribution is a known, common probability distribution. The posterior distribtion is proper because multiplying the kernel (shown in **Part A)** by the normalizing constant ensures its density integrates to 1 over the entire parameter space, $\Theta$.

#### **Question:** Check that this is not circular reasoning.

## Part C

*Now, suppose you are measuring the number of seconds between lightning strikes during a storm, your prior is* $Gamma(0.1,1.0)$*, and your data is* $$(x_1,\dotsc,x_8) = (20.9, 69.7, 3.6, 21.8, 21.4, 0.4, 6.7, 10.0).$$ *Plot the prior and posterior p.d.f.s. (Be sure to make your plots on a scale that allows you to clearly see the important features.)*

```{r}
# define thetas (updated based on resulting scale)
thetas<-seq(0.001, 0.15, length.out = 1000)
# define data
data <- c(20.9, 69.7, 3.6, 21.8, 21.4, 0.4, 6.7, 10.0)
# define gamma distribution (prior)
a <- 0.1
b <-1.0
prior <- dgamma(x = thetas, shape = a, rate = b)
# define the posterior dist
n <- length(data) # define length of data
posterior <- dgamma(x = thetas, shape = a+n, rate = b + sum(data))

# plot prior
plot(thetas, prior,type = "l", col = "blue",lwd=2,
     xlab = "Simulated numbers of seconds between lightning strikes",ylab = "Density", yaxt = "n", xlim = range(thetas), yaxt="none")

# overlay posterior
par(new = TRUE)
# plot posterior
plot(thetas, posterior,type = "l", col = "red", lty = 3, lwd=2,
     axes = FALSE, xlab = "", ylab = "",xlim = range(thetas),yaxt="none")
# add legend
legend("topright",legend = c("Prior", "Posterior"),
       col = c("blue", "red"), lty = c(1, 2))

```

## Part D

*Give a specific example of an application where an Exponential model would be reasonable. Give an example where an Exponential model would NOT be appropriate, and explain why.*

**Reasonable:** The Exponential model is often used to model the time between events. Thus, an appropriate application may be to model the wait time between customers served at a store, assuming a constant rate and independence of events, as well as the probability of the next event occurring not depending on how much time has already elapsed.

**Unreasonable:** One example where an exponential model would not be appropriate is for modeling coin flips. A coin flip has discrete, binary outcomes and exponential distributions model continuous outcomes.

## Question 3

*Priors, Posteriors, Predictive Distributions (Hoff, 3.9) An unknown quantity* $Y | \theta$ *has a Galenshore(*$a, \theta$*) distribution if its density is given by* $$p(y | \theta) = \frac{2}{\Gamma(a)} \; \theta^{2a} y^{2a - 1} e^{-\theta^2 y^2}$$ *for* $y>0, \theta >0, a>0.$ *Assume for now that* $a$ *is known and* $\theta$ *is unknown and a random variable. For this density,* $$E[Y] = \frac{\Gamma(a +1/2)}{\theta \Gamma(a)}$$ *and* $$E[Y^2] = \frac{a}{\theta^2}.$$

### Part A

*Identify a class of conjugate prior densities for* $\theta$*. Assume the prior parameters are* $c$ a*nd* $d$*, which are fixed and known. That is, state the prior distribution for* $\theta,$ *which will have known and fixed parameters* $c, d$ *such that the resulting posterior is conjugate. Plot a few members of this class of densities.*

**Suppose** $\theta$ **also has Galenshore prior density, where** $\theta \sim \text{Galenshore}(c,d)$**.** The prior parameters are $c$ and $d$, which are fixed, known, and greater than 0. So, we have

$$
p(\theta | c,d) = \frac{2}{\Gamma(c)} \; d^{2c} \theta^{2c - 1} e^{-d^2 \theta^2}\qquad \theta>0
$$

. We will show that the posterior density, $p(\theta|y_{1:n})$, is conjugate (i.e., the posterior takes on an updated form of the prior). Let $y_1, ..., y_n$ be independent observations such that $Y_i|\theta \sim \text{Galenshore}(a,\theta)$. By Bayes' Theorem,

$$
p(\theta|y_{1:n}) = \frac{p(\theta, y_{1:n})}{p(y_{1:n})} = \frac{p(\theta|y_{1:n})p(\theta)}{p(y_{1:n})}
$$. Since we are conditioning on the data, it is a constant, so we have

$$
\frac{p(\theta|y_{1:n})p(\theta)}{p(y_{1:n})} \propto p(\theta|y_{1:n})p(\theta)
$$. Then, substituting in our likelihood and prior gives the following:

$$
p(\theta|y_{1:n})p(\theta) = \prod_{i=1}^n(\frac{2}{\Gamma(a)} \; \theta^{2a} y_i^{2a - 1} e^{-\theta^2 y_i^2})*(\frac{2}{\Gamma(c)} \; d^{2c} \theta^{2c - 1} e^{-d^2 \theta^2})
$$

. Combining terms and absorbing terms which do not depend on $\theta$ into the constant of proportionality yields

$$
\begin{aligned}
p(\theta \mid y_{1:n})&\propto\theta^{2an}\exp\left(-\theta^2\sum_{i=1}^n y_i^2\right)\theta^{2c-1}\exp(-d^2\theta^2) = \theta^{2(c+na)-1}\exp\left[-\theta^2\left(d^2+\sum_{i=1}^n y_i^2\right)\right].
\end{aligned}
$$

. Therefore, we found $p(\theta|y_{1:n}) \propto \text{Galenshore}(\theta| c' = c + na, d' = \sqrt{d^2+\sum_{i=1}^n y_i^2})$, so the resulting posterior is conjugate.

Here, we plot a few members of this class of densities:

```{r}
# Galenshroe prior density function
# inputs: c,d, thetas
# output: priod density for each theta
galenshore_prior <- function(c, d, thetas) {
  # galenshore density
  2 / gamma(c) * d^(2 * c) * thetas^(2 * c - 1) * exp(-d^2 * thetas^2)
}
# create a grid of theta values
thetas <- seq(0.001, 3, length.out = 1000)

# deifne some different c,d values for comparisons of diff members of class
c1 <- 1
d1 <- 1

c2 <- 2
d2 <- 1

c3 <- 3
d3 <- 3

# compute prior densities usign function above
prior1 <- galenshore_prior(c1, d1, thetas)
prior2 <- galenshore_prior(c2, d2, thetas)
prior3 <- galenshore_prior(c3, d3, thetas)

# plot priors and overlay
plot(thetas, prior1, type = "l", col = "blue", lwd = 2, xlab = expression(theta),
     ylab = "Prior density",main = "Galenshore Prior Comparison",
  ylim = c(0, max(prior1, prior2, prior3)))

lines(thetas, prior2, col = "red", lwd = 2, lty = 2)
lines(thetas, prior3, col = "green", lwd = 2, lty = 3)

legend("topright", legend = c("c = 1, d = 1", "c = 2, d = 1", "c = 3, d = 3"),
  col = c("blue", "red", "green"), lty = c(1, 2, 3),lwd = 2)
```

Here, I used ChatGPT to help me fix formatting for long equations and to help me plot the densities.

### Part B

*Let* $Y_1, \ldots, Y_n \stackrel{iid}{\sim}$ *Galenshore(*$a, \theta$*). Find the posterior distribution of* $\theta | y_{1:n}$ *using a prior from your conjugate class.*

Let $\theta \sim \text{Galenshore}(c=1,d=1)$. By the proof-sketch shown in **Part A**, we have that

$\theta \mid y_{1:n} \sim \text{Galenshore}(1+na,\sqrt{1^2+\sum_{i=1}^{n}y_i^2})$, so

$$p(\theta \mid y_{1:n}) =\frac{2}{\Gamma(1+na)} \; (1+\sum_{i=1}^ny_i^2)^{1+na} \theta^{2(1+na) - 1} e^{-(1+\sum_{i=1}^ny_i^2) \theta^2}\qquad \theta>0$$.

### Part C

*Show that* $$\frac{p(\theta_a | y_{1:n})}{p(\theta_b | y_{1:n})} = \bigg( \frac{\theta_a}{\theta_b} \bigg)^{2(an + c) - 1}
e^{(\theta_b^2 - \theta_a^2)(d^2 + \sum y_i^2)},$$ *where* $$\theta_a, \theta_b \sim \text{Galenshore}(c,d).$$ *Identify a sufficient statistic.*

Let\
$$\theta_a, \theta_b \sim \text{Galenshore}(c,d).$$ From **Part B**, we know

$$p(\theta_a \mid y_{1:n}) =\frac{2}{\Gamma(1+na)} \; (1+\sum_{i=1}^ny_i^2)^{1+na} \theta_a^{2(1+na) - 1} e^{-(1+\sum_{i=1}^ny_i^2) \theta_a^2}\qquad \theta_a>0$$

and $$p(\theta_b \mid y_{1:n}) =\frac{2}{\Gamma(1+na)} \; (1+\sum_{i=1}^ny_i^2)^{1+na} \theta_b^{2(1+na) - 1} e^{-(1+\sum_{i=1}^ny_i^2) \theta_b^2}\qquad \theta_b>0$$. Dividing $$\frac{p(\theta_a | y_{1:n})}{p(\theta_b | y_{1:n})}$$ and canceling out like terms (that don't depend on $\theta_a$ or $\theta_b$) gives $$\frac{p(\theta_a | y_{1:n})}{p(\theta_b | y_{1:n})} = \bigg( \frac{\theta_a}{\theta_b} \bigg)^{2(an + c) - 1}
e^{(\theta_b^2 - \theta_a^2)(d^2 + \sum_{i=1}^n y_i^2)}$$. A sufficient statistic is $T(y_{1:n}) = \sum_{i=1}^ny_i^2$.

#### Question: Do I have to defend sufficient stat?

### Part D

*Determine* $E[\theta | y_{1:n}]$*.*

In the introduction to Question 3, we are given\
$$E[Y] = \frac{\Gamma(a +1/2)}{\theta \Gamma(a)}$$

for the Galenshore distribution. Since the posterior is also Galenshore with updated parameters, $$
\theta \mid y_{1:n} \sim \operatorname{Galenshore}\left( c+na,; \sqrt{d^2+\sum_{i=1}^n y_i^2} \right),
$$we have

$$\boxed{E[\theta | y_{1:n}] = \frac{\Gamma((an+c) +1/2)}{\sqrt{(d^2+\sum_{i=1}^ny_i^2)} \Gamma(an+c)}}.$$

### Part E

*Show that the form of the posterior predictive density is* $$p(y_{n+1} | y_{1:n}) =  \frac{2 y_{n+1}^{2a - 1} \Gamma(an + a + c)}{\Gamma(a)\Gamma(an + c)}
\frac{(d^2 + \sum y_i^2)^{an + c}}{(d^2 + \sum y_i^2 + y_{n+1}^2)^{(an + a + c)}}.$$

To calculate the posterior predictive density, we use

$$
p(y_{n+1} \mid y_{1:n})
=
\int_0^\infty
p(y_{n+1} \mid \theta)
p(\theta \mid y_{1:n})
\, d\theta.
$$

The likelihood for the new observation is

$$
p(y_{n+1} \mid \theta)
=
\frac{2}{\Gamma(a)}
\theta^{2a}
y_{n+1}^{2a-1}
\exp\left(-\theta^2 y_{n+1}^2\right).
$$

From Part A, the posterior density is

$$
p(\theta \mid y_{1:n})
=
\frac{2}{\Gamma(an+c)}
\left(d^2+\sum_{i=1}^n y_i^2\right)^{an+c}
\theta^{2(an+c)-1}
\exp\left[
-\theta^2
\left(d^2+\sum_{i=1}^n y_i^2\right)
\right].
$$

Substituting the likelihood and posterior into the posterior predictive integral gives

$$
\begin{aligned}
p(y_{n+1} \mid y_{1:n})
&=
\int_0^\infty
\left[
\frac{2}{\Gamma(a)}
\theta^{2a}
y_{n+1}^{2a-1}
\exp\left(-\theta^2 y_{n+1}^2\right)
\right] \\
&\quad \times
\left[
\frac{2}{\Gamma(an+c)}
\left(d^2+\sum_{i=1}^n y_i^2\right)^{an+c}
\theta^{2(an+c)-1}
\exp\left(
-\theta^2
\left(d^2+\sum_{i=1}^n y_i^2\right)
\right)
\right]
\, d\theta \\
&=
\frac{4y_{n+1}^{2a-1}}
{\Gamma(a)\Gamma(an+c)}
\left(d^2+\sum_{i=1}^n y_i^2\right)^{an+c} \\
&\quad \times
\int_0^\infty
\theta^{2(an+a+c)-1}
\exp\left[
-\theta^2
\left(
d^2+\sum_{i=1}^n y_i^2+y_{n+1}^2
\right)
\right]
\, d\theta.
\end{aligned}
$$

To obtain a proper posterior (integrates to 1 over the parameter space, $\Theta$, we multiply and divide by $$ \frac{2}{\Gamma{(an+a+c)}} (d^2+\sum_{i=1}^ny_i^2+y_{n+1}^2)$$, which gives

we obtain

$$
\boxed{
p(y_{n+1} \mid y_{1:n})
=
\frac{
2y_{n+1}^{2a-1}\Gamma(an+a+c)
}{
\Gamma(a)\Gamma(an+c)
}
\frac{
\left(d^2+\sum_{i=1}^n y_i^2\right)^{an+c}
}{
\left(
d^2+\sum_{i=1}^n y_i^2+y_{n+1}^2
\right)^{an+a+c}
}.
}
$$

I used ChatGPT to re-format my long equations. Algebra is my own.
